Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-10-21 paper-bedpca.tex
save

For example, we show that PC18 to PC40 in the UK Biobank are capturing LD structure. Using our automatic algorithm for removing long-range LD region, we are able to recover only 17 PCs that capture population structure only. Therefore, we recommend using only 17 PCs from the UK Biobank. Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16. We provide a fast way to project new individuals that is mostly free of this shrinkage bias. We also show how to restrict to individuals that are genetically homogeneous based on PCA.

@@

In this paper, for many applications, we use the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}. We reexport this function in R package bigutilsr for convenience.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}. 
To summarise the contribution of each variant in all PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of PC loadings.  
This captures both population-differentiating variants as well as long-range LD regions. 
Then, to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

@@

@@

@@

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Sat Oct 19 22:53:58 2019
\begin{table}[ht]
\centering
\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
  \hline
 & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black & NA \\ 
  \hline
AFR\_ACB &  &  &  &  &  &  &  &  &  & 1687 & 580 & 28 & 252 \\ 
  AFR\_ASW &  &  &  &  &  &  &  &  &  & 365 & 12 & 12 & 41 \\ 
  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  & 16 \\ 
  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 12 &  & 5 \\ 
  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 33 &  & 6 \\ 
  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 23 &  & 2 \\ 
  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 403 & 3 & 66 \\ 
   \hline
AMR\_CLM & 1 &  &  & 47 &  &  &  &  &  &  &  &  & 127 \\ 
  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 21 \\ 
  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  & 14 \\ 
   \hline
EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  &  \\ 
  EAS\_CHB & 1 &  &  &  &  &  &  & 369 & 10 &  &  &  & 9 \\ 
  EAS\_CHS &  &  &  &  &  &  &  & 398 & 8 &  &  &  & 18 \\ 
  EAS\_JPT &  &  &  &  &  &  &  & 2 & 23 &  &  &  & 114 \\ 
  EAS\_KHV &  &  &  &  &  &  &  & 3 & 14 &  &  &  & 19 \\ 
   \hline
EUR\_CEU & 323808 & 6547 & 348 & 4904 &  &  &  & 1 &  &  &  &  & 1358 \\ 
  EUR\_FIN & 3 &  &  & 115 &  &  &  &  &  &  &  &  & 1 \\ 
  EUR\_GBR & 62059 & 3682 & 79 & 482 &  &  &  &  &  &  & 1 &  & 268 \\ 
  EUR\_IBS & 35 &  & 5 & 494 &  &  &  &  &  &  &  &  & 10 \\ 
  EUR\_TSI & 20 &  & 1 & 156 &  &  &  &  &  &  &  &  & 3 \\ 
   \hline
SAS\_BEB &  &  &  &  & 10 &  & 41 &  &  &  &  &  & 4 \\ 
  SAS\_GIH &  &  &  &  & 292 &  &  &  &  &  &  &  & 3 \\ 
  SAS\_ITU & 1 &  &  &  & 270 & 1 &  &  & 224 & 2 &  &  & 69 \\ 
  SAS\_PJL &  &  &  &  & 1406 & 641 &  &  & 82 & 2 &  &  & 147 \\ 
  SAS\_STU &  &  &  &  & 39 &  &  &  & 198 &  &  &  & 47 \\ 
   \hline
NA & 45101 & 2526 & 111 & 9614 & 3699 & 1106 & 180 & 730 & 1188 & 2237 & 2037 & 75 & 7051 \\ 
   \hline
  \end{tabular}
\end{table}

@@

\noindent[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]

@@

>>>>

For example, we show that PC18 to PC40 in the UK Biobank are capturing LD structure. Using our automatic algorithm for removing long-range LD region, we are able to recover only 17 PCs that capture population structure only. Therefore, we recommend using only 17 PCs from the UK Biobank. Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16. We provide a fast way to project new individuals and that is not affected by this shrinkage bias. We also show how to restrict to individuals that are genetically homogeneous based on PCA.

@@

In this paper, for many applications, we use the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}. We reexport function \texttt{covRob} in R package bigutilsr for convenience.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}.
We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
Then, we compute $K$ PC scores and loadings.
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

@@

We remove the detected outlier variants, compute the PC scores and loadings again, and iterate until there is no detected outlier variant anymore.
This procedure is implemented in function \texttt{bed\_autoSVD} or R package bigsnpr.

@@

\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL]

\noindent[TALK ABOUT EFFECT OF MISSING VALUES ON PCA?]

\noindent[TALK ABOUT REMOVING OF LRLDR?]

\subsection{Performance}

To show that we provide very performant implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
It takes 22 minutes to perform a first phase of clumping on 488,371 individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 computation of PCs and 5 outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

It takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic outlier detection technique. It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data to the PCA computed using the 1000G data. Once projections are available, et takes only a couple minutes to perform the ancestry estimation.

[TALK ABOUT KNN]

@@

% latex table generated in R 3.6.0 by xtable 1.8-4 package
% Mon Oct 21 15:14:34 2019
\begin{table}[ht]
\centering
\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method. See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.} 
\label{tab:ancestry-pred}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
  \hline
 & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black & NA \\ 
  \hline
AFR\_ACB &  &  &  &  &  &  &  &  &  & 1832 & 580 & 31 & 259 \\ 
  AFR\_ASW &  &  &  &  &  &  &  &  &  & 315 & 13 & 10 & 39 \\ 
  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 155 & 1 & 29 \\ 
  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 20 &  & 4 \\ 
  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 34 &  & 4 \\ 
  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 26 &  &  \\ 
  AFR\_YRI &  &  &  &  &  &  &  &  &  & 5 & 469 & 3 & 81 \\ 
   \hline
AMR\_CLM & 1 &  &  & 50 &  &  &  &  &  &  &  &  & 142 \\ 
  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 20 \\ 
  AMR\_PEL &  &  &  & 2 &  &  &  &  &  &  &  &  & 19 \\ 
   \hline
EAS\_CDX &  &  &  &  &  &  &  & 2 &  &  &  &  & 1 \\ 
  EAS\_CHB & 1 &  &  &  &  &  &  & 445 & 10 &  &  &  & 7 \\ 
  EAS\_CHS &  &  &  &  &  &  &  & 380 & 9 &  &  &  & 21 \\ 
  EAS\_JPT &  &  &  &  &  &  &  & 3 & 25 &  &  &  & 122 \\ 
  EAS\_KHV &  &  &  &  &  &  &  & 4 & 16 &  &  &  & 22 \\ 
   \hline
EUR\_CEU & 325725 & 6675 & 350 & 4855 &  &  &  & 1 &  &  &  &  & 1363 \\ 
  EUR\_FIN & 2 &  &  & 118 &  &  &  &  &  &  &  &  & 1 \\ 
  EUR\_GBR & 57497 & 3486 & 79 & 454 &  &  &  &  &  &  & 1 &  & 244 \\ 
  EUR\_IBS & 28 &  & 4 & 452 &  &  &  &  &  &  &  &  & 8 \\ 
  EUR\_TSI & 25 &  & 1 & 177 &  &  &  &  &  &  &  &  & 5 \\ 
   \hline
SAS\_BEB &  &  &  &  & 11 &  & 57 &  &  &  &  &  & 7 \\ 
  SAS\_GIH &  &  &  &  & 334 &  &  &  & 1 &  &  &  & 4 \\ 
  SAS\_ITU & 1 &  &  &  & 314 & 1 &  &  & 264 & 3 &  &  & 82 \\ 
  SAS\_PJL & 1 &  &  &  & 1749 & 767 &  &  & 104 & 2 &  &  & 177 \\ 
  SAS\_STU &  &  &  &  & 38 &  &  &  & 185 &  &  &  & 40 \\ 
   \hline
NA & 47748 & 2594 & 110 & 9705 & 3270 & 980 & 164 & 669 & 1133 & 2140 & 1906 & 73 & 6970 \\ 
   \hline
\end{tabular}
\end{table}

@@

\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP]

@@

\noindent[DISCUSS LACK OF CERTAIN POPULATIONS IN 1000G]
